---
title: "Assignment 2"
author: "Neriman Gürsoy"
date: "11/12/2020"
output: 
  html_document:
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
---
<style>

body {
  color: #708090;
  font-family: Calibri;
  background-color: #F8F8FF;
}
</style>


## IMPLEMENTATION OF THE dployr and ggplot2 PACKAGES
## ELECTRICITY MARKET PRICES REPORT

This electricity report shows that implementation and visualization by using dplyr and ggplot2 packages for 01-30 September 2020 EXIST dataset. 

## 1.About Energy Exchange Istanbul (EXIST)

Energy Exchange Istanbul (EXIST) was officially established on March 12, 2015 upon the registration processes completed subject to the provisions of the Electricity Market Law of 14.03.2013 numbered 6446 and  Turkish Trade Law of 12 Mart 2015 numbered 6102.
Main objective and principal business activity is to plan, establish, develop, and manage energy market in a transparent manner that fulfills the requirements of energy market. Energy market in Turkey changes hourly and the related data can be found in the official webpage of EXIST. You can [click](https://rapor.epias.com.tr/rapor/xhtml/ptfSmfListeleme.xhtml) and filter the date you wish to check hourly data.

## 2. About the Dataset

This dataset consists of six different values:

  1. Date
  2. Market clearing price (MCP)
  3. System Marginal Price (SMP)
  4. Positive Imbalance Price
  5. Negative Imbalance Price
  6. System Marginal Price Direction
  
Before we get into the details of the report, there are a few terms about the dataset that we should mention. These are **energy deficit** and **energy surplus**. If there is a system-wide need for extra electricity production (i.e. actual demand > predicted demand), then it is called an energy Deficit. If the situation is the opposite (i.e. actual demand < predicted demand), then it is called an Energy Surplus. System Marginal Price(SMP) is always higher than Market Clearing Price(MCP) if system has Energy Deficit, and lower if there is Energy Surplus. Market operator also penalizes the operations in BPM by 3%. This is called **Imbalance Price**. Negative (Deficit) Imbalance Price is calculated as max(MCP,SMP)1.03 and Positive Imbalance Price is calculated as min(MCP,SMP)0.97.

## 3. Analysis of the Dataset

After downloading the data between 1 September 2020 and 30 September 2020 from the official webpage of EXIST, the analyses can be started. First of all lubridate, tidyverse and ggplot packages which are used in the report should be loaded. Then we can upload the date which is csv file format.

```{r setup, include= FALSE}
knitr::opts_chunk$set(fig.align="center")

```


```{r , message=FALSE, warning=FALSE}

library (readr)
library (dplyr)
library (tidyverse)
library (lubridate)
library (ggplot2)

Existdata <- read.csv("ptf-smf.csv")

```

We can use the glimpse function to inspect our data. By using it,each column is represented in a row with its data type and first few entries. We have 720 rows and 6 variables.

```{r}
Existdata %>% glimpse()
```

### 3.1. Rename columns and rearrange the data frame

The *date* column is not date type, to convert from char to datetime.

```{r , message= FALSE}

Existdata$Date <- gsub(pattern = "\\.","-",Existdata$Date)
Existdata$Date <- as.POSIXct(Existdata$Date,format = "%d-%m-%y %H:%M")

Existdata %>% glimpse()

```

Renaming of the columns by using transmute() function in order to have a better comprehension.


```{r, message= FALSE, warning= FALSE}

raw_df <- Existdata %>% transmute(Date,
                                  MCP = MCP, 
                                  SMP = SMP, 
                                  PositiveIP = Positive.Imbalance.Price..TL.MWh., 
                                  NegativeIP = Negative.Imbalance.Price..TL.MWh., 
                                  SMPDirection = SMP.Direction )

raw_df %>% glimpse()
```


### 3.2. Market clearing price (MCP)

**3.2.1. Hourly MPC**

Since the electricity energy market prices are planned on an hourly basis, it will be useful to find the hourly average, minimum, and maximum values in order to gain some insights about the data. 

```{r, message= FALSE}
plot1 <- raw_df %>% group_by(Hour=lubridate::hour(Date)) %>% 
  summarize(AvgMCP_Hour=mean(MCP), MinMCP_Hour=min(MCP), MaxMCP_Hour=max(MCP) ) %>% 
  print(n=5)
```


```{r}
 ggplot(plot1, aes(x=Hour)) +
    geom_line(aes(y = MinMCP_Hour, color = "Minimum hourly MCP"))+
    geom_line(aes(y = MaxMCP_Hour, color = "Maximum hourly MCP"))+
    geom_line(aes(y = AvgMCP_Hour, color = "Average hourly MCP"))+
    labs( title = "Average, Minimum and Maximum Hourly MCP Values",
          subtitle= "Energy Exchange Turkey (EXIST), between September 01 and 30",
          color = "MCP",
          x = "Hours",
          y = " MCP Values")+
    theme_minimal()
```

**3.2.2. Daily  MPC**

```{r, message=FALSE}
plot2 <- raw_df %>% group_by(Day=lubridate::day(Date)) %>% 
  summarize(AvgMCP_day=mean(MCP), MinMCP_day=min(MCP), MaxMCP_day=max(MCP) ) %>% 
  print(n=5)
```

```{r}

 ggplot(plot2, aes(x=Day)) +
    geom_line(aes(y = MinMCP_day, color = "Minimum hourly MCP"))+
    geom_line(aes(y = MaxMCP_day, color = "Maximum hourly MCP"))+
    geom_line(aes(y = AvgMCP_day, color = "Average hourly MCP"))+
    labs( title = "Average, Minimum and Maximum Daily MCP Values",
          subtitle= "Energy Exchange Turkey (EXIST), between September 01 and 30",
          color = "MCP",
          x = "Days",
          y = " MCP Values")+
    theme_minimal()
```

**3.2.3. Weekly  MPC**

Weekly MPC prices for the minimum, maximum, and average values can be seen below. Week numbers correspond to the sequence in a year of 52 weeks.

```{r, message= FALSE}
plot3 <- raw_df %>% group_by(Week=lubridate::week(Date)) %>% 
  summarize(AvgMCP_week=mean(MCP), MinMCP_week=min(MCP), MaxMCP_week=max(MCP) ) %>% 
  print(n=5)
```

```{r}
 ggplot(plot3, aes(x=Week)) +
    geom_line(aes(y = MinMCP_week, color = "Minimum hourly MCP"))+
    geom_line(aes(y = MaxMCP_week, color = "Maximum hourly MCP"))+
    geom_line(aes(y = AvgMCP_week, color = "Average hourly MCP"))+
    labs( title = "Average, Minimum and Maximum Weekly MCP Values",
          subtitle= "Energy Exchange Turkey (EXIST), between September 01 and 30",
          color = "MCP",
          x = "Weeks",
          y = " MCP Values")+
    theme_minimal()
```

### 3.3. System Marginal Price (SMP)

**3.3.1. Daily change of the SMP**

```{r, message= FALSE}
plot4 <- raw_df %>% group_by(Day=lubridate::day(Date)) %>% 
  summarize(AvgSMP_day=mean(SMP), MinSMP_day=min(SMP), MaxSMP_day=max(SMP)) %>%
  ungroup()
  plot4

```

```{r}
    ggplot(plot4, aes(x=Day, color = "Days"))+
    geom_line(aes(y = AvgSMP_day, color = "Average daily SMP")) +
    geom_line(aes(y = MinSMP_day, color = "Minimum daily SMP")) +
    geom_line(aes(y = MaxSMP_day, color = "Maximum daily SMP")) +
    theme_minimal() +
    labs(title = "Minimum, Maximum and Average Price of Daily SMP",
         subtitle = "Energy Exchange Turkey (EXIST), between September 01 and 30",
         x= "Days",
         y= "Prices",
         color = "Prices")
```


### 3.4.Relationship between MCP and SMP 

We can the following plot in order to obtain daily average of the MCP and SMP.

```{r, message=FALSE}
plot5 <- raw_df %>%
        group_by(Day = lubridate::day(Date))%>%
        summarise(Average_MCP = mean(MCP),
                  Average_SMP = mean(SMP),
                  Difference = abs(mean(MCP) - mean(SMP))) %>%
        ungroup()%>%
        arrange(desc(Difference))
        
knitr::kable(plot5, format="markdown")
```

**Graph of the daily average value of MCP and SMP with line graph:**

```{r}
  ggplot(plot5, aes(x = Day)) +
    geom_line(aes(y = Average_MCP, color = "Daily Average MCP")) +
    geom_line(aes(y = Average_SMP, color = "Daily Average SMP")) +
    theme_minimal() +
    labs(title = "Comparison of daily SMP and MCP",
         x="Days",
         y="Price",
         color="Price Type")

```
```{r}
  ggplot(plot5, aes(x=Day)) + 
    geom_line(aes(y = Difference, color = "Difference of average values")) +
    labs(x = "Day", y = "TL/MWh",
       title = "Difference between Daily average MCP-SMP Change Graph",
       subtitle = "Energy Exchange of Turkey - EXIST, between September 01 and September 31 ")+
  theme_minimal() +
  theme(legend.position = "center")
```

The graph shows that the highest difference between daily average value of SMP and MPC is observed in the third day.


**Graph of the weekly average value of MCP and SCP with bar graph:**

```{r, message= FALSE}
  plot6 <-raw_df %>%
  group_by(Week = lubridate::week(Date)) %>%
  summarise(Weekly_MCP = mean(MCP),
            Weekly_SMP = mean(SMP),
            Difference = abs(mean(MCP) - mean(SMP))) %>%
  ungroup() %>%
  arrange(desc(Difference))
plot6
 
```

```{r}
 ggplot(plot6,aes(x = Week, y = Difference, fill = "Weeks")) +
    geom_bar(stat = "identity") +
    theme_minimal()+
    theme(legend.position = "none") +
    labs(title = "Difference Between Weekly MCP and SMP ",
         x = "Week",
         y = "Average Price Difference")
```

**Scatter Graph of the relationship between MCP and SMP values:**

The plot below shows that for most of the days, MCP lies 200-400 and SMP lies 100-500. Also we can say that there is an outlier value which SMP is equal 2000.

```{r, message= FALSE, warning= FALSE}

  ggplot(raw_df, aes(x=MCP, y=SMP, color=Date)) + 
  geom_point() + 
  labs(x="MCP", y="SMP",  title="MCP and SMP Prices in September 2020") + 
  theme_get() + 
  geom_abline(slope=1, intercept=0, color="red")

```


### 3.5. Negative and positive imbalance

3.5.1. Negative and positive imbalance price in first day

Negative and Positive imbalance prices can be seen below for the first day of the September, 2020.

```{r}
raw_df %>%
  select(Date, PositiveIP, NegativeIP)%>%
  group_by(Day = lubridate::day(Date), Hour = lubridate::hour(Date))%>%
  select(Day,Hour, PositiveIP, NegativeIP)%>%
  filter(Day==1)%>%
  
  ggplot(aes(Hour, color = "Hours"))+
    geom_line(aes(y=PositiveIP, color = "Positive imbalance price")) +
    geom_line(aes(y=NegativeIP, color = "Negative imbalance price")) +
    labs(x = "Hours", y = "TL/MWh",
         title= "Pozitive and Negative Imbalance Price of First Day",
         color = "Imbalance Price") +
    theme_minimal()

```

3.5.2. Negative and positive imbalance price in first day

Negative and Positive imbalance prices can be seen below for the last day of the September, 2020.

```{r}
raw_df %>%
  select(Date, PositiveIP, NegativeIP)%>%
  group_by(Day = lubridate::day(Date), Hour = lubridate::hour(Date))%>%
  select(Day,Hour, PositiveIP, NegativeIP)%>%
  filter(Day==30)%>%
  
  ggplot(aes(Hour, color = "Hours"))+
    geom_line(aes(y=PositiveIP, color = "Positive imbalance price")) +
    geom_line(aes(y=NegativeIP, color = "Negative imbalance price")) +
    labs(x = "Hours", y = "TL/MWh",
         title= "Pozitive and Negative Imbalance Price of Last Day",
         color = "Imbalance Price") +
    theme_minimal()
```

3.5.3. Graph of daily difference of negative and positive imbalance price

```{r}

  plot7<-raw_df %>%
        select(Date, PositiveIP, NegativeIP)%>%
        group_by(Day = lubridate::day(Date), Hour = lubridate::hour(Date))%>%
        select(Day,Hour, PositiveIP, NegativeIP)%>%
        transmute(Day,Hour,PositiveIP, NegativeIP, difference=NegativeIP-PositiveIP)%>%
        ungroup()%>%
        arrange(desc(difference)) %>%
        print(n=5)

```

### 3.6. SMP Direction

These are **energy deficit**, **energy surplus**, and **energy balance** due to the relationship between actual demand and predicted demand. In the data we examine, there are energy deficit, surplus and balance states on an hourly basis. This information is given in the column called SMP Direction.

```{r, message= FALSE}
plot8 <- raw_df %>% group_by(SMPDirection) %>% summarize(count=n())%>%print()
```
```{r}
ggplot(plot8, aes(x="", y=count, fill=SMPDirection)) + geom_bar(stat="identity", width=1) + 
  coord_polar("y") + 
  theme_minimal() +
  labs( title="Energy Imbalance in September 2020",
        fill="SMP Direction")
```

3.6.1. Hourly Energy Deficit, Energy Surplus,and Balance Distribution

```{r, message = FALSE}
plot9 <- raw_df %>% group_by(Hour=lubridate::hour(Date)) %>%
         summarise(Surplus=sum(MCP>SMP), Deficit=sum(MCP<SMP), Balance=sum(MCP==SMP)) %>% 
         ungroup() %>%
         select(Hour, Surplus, Deficit, Balance) %>% 
         print(n=5)

```

```{r}
  ggplot(plot9, aes(x = Hour, color = "Hours")) +
    geom_line(aes(y = Surplus, color = "Number of surplus")) +
    geom_line(aes(y = Deficit, color = "Number of deficit")) +
    geom_line(aes(y = Balance, color = "Number of balance")) +
    
  labs(x = "Hour",
       y = "Numbers",
       title = "Hourly Energy Imbalance",
       color = "SMP Direction Type") +
  theme_minimal()
```

This graph shows that, especially between 8 and 23 pm, the number of deficit is greater than both number of surplus and balance. The deficit means that, the energy demand is more than the predicted energy. That is the more energy used than the predicted.According to the daily analysis, in each hour, we have more deficit value than surplus and balance. That is, we use more energy than predict.


3.6.2. Graph of the Hourly Deficit Energy Usage 

```{r, message= FALSE}
plot10 <- raw_df %>%
        group_by(Hours = lubridate::hour(Date))%>%
        transmute(EnergyUsage = ifelse(SMPDirection == "?Energy Surplus", "Surplus", "Deficit")) %>%
        group_by(EnergyUsage, Hours) %>%
        filter(EnergyUsage=="Deficit") %>%
        summarise(Number = n())%>% 
        print(n=5)
  

```

```{r}
  ggplot(plot10, aes(x = Hours, y = Number, fill = Number)) +
         geom_bar(stat = "identity") +
         theme_minimal() +
         labs(title = "Hourly Deficit Energy Usage ",
         fill = "Number of Energy Usage",
         x="Hours",
         y = "Count")
```

This bar chart shows the hourly number of deficit energy usage. This result shows that the more deficit energy usage observed in the 6 pm.In
